Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method

### 2023-08-15 manu-draft.qmd
added info about synthetic time series

For @eq-ebase, the terms on the right side of the equation are gross primary production (P, $aPAR$), ecosystem respiration (R), and gas exchange (D, remainder), respectively, processes that result in the change in dissolved oxygen ($dC$) per unit time ($dt$); $H$ is the water depth. Production is modeled linearly as a function of surface irradiance (PAR, $\text{W}/\text{m}^2$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$). Respiration is modeled with the single parameter $R$ ($\text{mmol}/\text{m}^2/\text{d}$). Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ ($\text{mmol}/\text{m}^3$) , and the saturation (or equilibrium) concentration, $C_{Sat}$ ($\text{mmol}/\text{m}^3$) calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92]. The gas transfer velocity, $k_w$ in @eq-kw, is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, $\text{m}^2/\text{s}^2$) and the Schmidt number ($Sc$), which is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen in water [$S_c$ is computed from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "tbl-ebaseiotab")`).

@@

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each. As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches. Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@caffrey2004; @Wanninkhof14] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space. Specifically, the prior mean for $R$ was chosen as 300 $\text{mmol}/\text{m}^2/\text{d}$ (\~10 $\text{g O}_2/\text{m}^2/\text{d}$) using a national synthesis of metabolic rates for US estuaries [Table 2 in @caffrey2004]. Although @caffrey2004 focused on small, shallow estuaries that are part of the NERRS, this dataset is the largest, longest, and most consistently collected in the US and it is reasonable to expect others may apply EBASE to these data. The prior mean for $a$ was chosen assuming that $P$ and $R$ are comparable at long time scales, such that $aPAR$ = $R$, or $a$ = $300/PAR$. Using the global mean surface $PAR$ of 70 $W/m^2$, the prior mean for $a$ is reasonably set at 4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following @Wanninkhof14. The prior standard deviations were chosen similarly based on values in @caffrey2004 using the standard deviation of the reported values. The standard deviation of $R$ was 171 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and the standard deviation for $a$ was set proportionally as 2.25 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, again assuming $P$ = $aPAR$ = $R$. The standard deviation for $b$ was set as 50% of the mean as an uninformed and reasonable value given that there are no comparable global or national estimates for the range of values. Together, the default priors for $a$, $R$, and $b$ are:

$$
a \sim N(4, 1)T(0,\infty) \;\; \left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)
$$ {#eq-aprior}

$$
R \sim N(300, 50)T(0,\infty) \;\; \text{mmol}/\text{m}^2/\text{d}
$$ {#eq-rprior}

$$
b \sim N(0.251, 0.1)T(0,0.502) \;\; \left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)
$$ {#eq-bprior}

where $N(\alpha, \beta)$ indicates a normal distribution with mean $\alpha$ and standard deviation $\beta$ and $T(\gamma, \theta)$ is 1 between $\gamma$ and $\theta$ and 0 elsewhere; $T$ has the effect of truncating the normal distribution to minimum and maximum values. All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase. Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ (\~twice the default mean). Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable parameter estimates. Examples of the prior distributions in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] are shown in @fig-priorplot.

@@

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $R$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs. These assessments evaluated if 1) EBASE estimates similar values for the known parameters in the synthetic time series, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from one year of continuous data from Apalachicola Bay, Florida. The input data were observations from the year 2021 at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15-minute observations of water temperature and salinity. Continuous meteorological observations from the nearby East Bay station (29.7694$^{\circ}$ N, --84.8815$^{\circ}$ W, \~10 km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR. Missing observations in the water quality and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st. A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a) and @eq-ebase as a forward calculating model with user-defined values for $a$, $R$, and $b$. All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series. The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14]. Accordingly, the synthetic dissolved oxygen time series and the observed water quality and meteorological data were used with EBASE to determine if reasonable estimates for the known parameters and metabolic rates were recovered.

@@

The prior distributions for the EBASE parameters follow truncated normal distributions with defined means ($\mu$) and standard deviations ($\sigma$)(Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the priors. Low and high values for the means and standard deviations were assessed following reasonable ranges from @caffrey2004 using logic similar to that for choosing the default priors. The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and proportionally for $a$ as 2.0 and 14.4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default mean values to test a range from high precision to completely uninformed values. The mean and standard deviation for the $b$ parameter were not tested and set as the default values for all analyses given that there is minimal prior knowledge on expectations for this parameter. The default prior distribution is already uninformed and we have no appropriate expectations on methods for varying this parameter. Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^4$ = 16 different combinations of prior distributions that were used with the synthetic time series in EBASE.

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $R$, and $b$). Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being one day and a maximum being the length of the time series. For a chosen optimization period, single estimates for $a$, $R$, and $b$ are returned by the model. A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if variation occurs at time scales less than the optimization period. Results from optimization periods of one, seven, and thirty days were compared for the synthetic one-year time series. Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times the three optimization periods) unique analyses for the assessment of changing priors of all parameters.

@@

![Synthetic time series for one year of continuous data at Apalachicola Bay. Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) synthetic $a$ parameter and (c) synthetic metabolic estimates for gross primary production (P), ecosystem respiration (R), and gas exchange (D). The $a$ parameter is shown as an areal value for comparison with the metabolic results. The $b$ parameter is also fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ and is not shown.](figs/synapa.png){#fig-synapa}

>>>>

For @eq-ebase, the terms on the right side of the equation are gross primary production ($P$, $aPAR$), ecosystem respiration ($R$), and gas exchange ($D$, remainder), respectively, processes that result in the change in dissolved oxygen ($dC$) per unit time ($dt$); $H$ is the water depth. Production is modeled linearly as a function of surface irradiance (PAR, $\text{W}/\text{m}^2$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$). Respiration is modeled with the single parameter $R$ ($\text{mmol}/\text{m}^2/\text{d}$). Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ ($\text{mmol}/\text{m}^3$) , and the saturation (or equilibrium) concentration, $C_{Sat}$ ($\text{mmol}/\text{m}^3$) calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92]. The gas transfer velocity, $k_w$ in @eq-kw, is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, $\text{m}^2/\text{s}^2$) and the Schmidt number ($Sc$), which is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen in water [$S_c$ is computed from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "tbl-ebaseiotab")`).

@@

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each. As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches. Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@caffrey2004; @Wanninkhof14] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space. Specifically, the prior mean for $R$ was chosen as 300 $\text{mmol}/\text{m}^2/\text{d}$ (\~10 $\text{g O}_2/\text{m}^2/\text{d}$) using a national synthesis of metabolic rates for US estuaries [Table 2 in @caffrey2004]. Although @caffrey2004 focused on small, shallow estuaries that are part of the NERRS, this dataset is the largest, longest, and most consistently collected in the US and it is reasonable to expect others may apply EBASE to these data. The prior mean for $a$ was chosen assuming that $P$ and $R$ are comparable at long time scales, such that $aPAR$ = $R$, or $a$ = $300/PAR$. Using the global mean surface $PAR$ of 70 $W/m^2$, the prior mean for $a$ is reasonably set at 4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following @Wanninkhof14. The prior standard deviations were chosen similarly based on values in @caffrey2004 using the standard deviation of the reported values. The standard deviation of $R$ was 170 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and the standard deviation for $a$ was set proportionally as 2.25 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, again assuming $aPAR$ = $R$. The standard deviation for $b$ was set as 50% of the mean as an uninformed and reasonable value given that there are no comparable global or national estimates for the range of values. Together, the default priors for $a$, $R$, and $b$ are:

$$
a \sim N(4, 2.25)I(0,\infty) \;\; \left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)
$$ {#eq-aprior}

$$
R \sim N(300, 170)I(0,\infty) \;\; \text{mmol}/\text{m}^2/\text{d}
$$ {#eq-rprior}

$$
b \sim N(0.251, 0.126)I(0,0.502) \;\; \left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)
$$ {#eq-bprior}

where $N(\alpha, \beta)$ indicates a normal distribution with mean $\alpha$ and standard deviation $\beta$ and $I(\gamma, \theta)$ is 1 between $\gamma$ and $\theta$ and 0 elsewhere; $I$ has the effect of truncating the normal distribution to minimum and maximum values. All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase. Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ (\~twice the default mean). Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable estimates for all parameters. Examples of the prior distributions in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] are shown in @fig-priorplot.

@@

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $R$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs. These assessments evaluated if 1) EBASE estimates similar values for the known parameters in the synthetic time series, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from one year of continuous data from Apalachicola Bay, Florida. The input data were observations from the year 2021 at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15-minute observations of water temperature and salinity. Continuous meteorological observations from the nearby East Bay station (29.7694$^{\circ}$ N, --84.8815$^{\circ}$ W, \~10 km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR. Missing observations in the water quality and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st.

A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]) and @eq-ebase as a forward calculating model with estimated values for $a$, $R$, and $b$. All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series. The time series was estimated using the observed data described above, the separate metabolic components and their parameters in @eq-ebase, and approximate metabolic rates from the Odum open-water method as implemented in the WtRegDO R package [@Beck21] as an approximate estimate for back-calculation of the EBASE parameters. In short, $P$, $R$, and $D$ as $\text{mmol}/\text{m}^2/\text{d}$ were estimated using the Odum method and the separate synthetic components for EBASE were created by estimating the corresponding parameters, such that $P /PAR = a$, $R$ as is, and $D = k_w\left(C_{sat} - C \right)$, where $k_w$ includes the $b$ parameter. Daily estimates for each were created as the daily average for each parameter fit with a polynomial spline to create a smoothed output. The synthetic dissolved oxygen time series was then created as the integrated sum of the separate components for $P$, $R$, and $D$ divided by water column depth. The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14]. Accordingly, the synthetic dissolved oxygen time series and the observed water quality and meteorological data were used with EBASE to determine if reasonable estimates for the known parameters and metabolic rates were recovered.

@@

The prior distributions for the EBASE parameters follow truncated normal distributions with defined means ($\mu$) and standard deviations ($\sigma$) (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the results. Low and high values for the means and standard deviations were assessed following reasonable ranges from @caffrey2004 using logic similar to that for choosing the default priors. The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and proportionally for $a$ as 2.0 and 14.4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default standard deviations to test a range from high precision to completely uninformed values. The mean and standard deviation for the $b$ parameter were not tested and set as the default values for all analyses given that there is minimal prior knowledge on expectations for this parameter. The default prior distribution is already uninformed and we have no appropriate expectations on methods for varying this parameter. Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^4$ = 16 different combinations of prior distributions that were used with the synthetic time series in EBASE.

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $R$, and $b$). Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being one day and a maximum being the length of the time series. For a chosen optimization period, single estimates for $a$, $R$, and $b$ are returned by the model. A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if variation occurs at time scales less than the optimization period. Results from optimization periods of one, seven, and thirty days were compared for the synthetic one-year time series. Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times three optimization periods) unique analyses for the assessment of changing priors of all parameters.

@@

![Synthetic time series for one year of continuous data at Apalachicola Bay. Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) synthetic $a$ parameter and (c) synthetic metabolic estimates for gross primary production ($P$), the negative of ecosystem respiration ($-R$), and gas exchange ($D$).](figs/synapa.png){#fig-synapa}
